What We Eval in the Shadows

### 2021-04-16 paper/main.tex
Side effects fixes

Contrary to the Javasscript \eval, which does not precise the scope in which to evaluate,the second argument of \eval in R, \c{envir}, specifically is dedicated ot that. 
The environment in which evaluation happens determines what is visible to the
computation started by \eval and the potential reach of its side-effects. The
second argument to \eval specifies that environment. Environments of \eval can be
classified into the following four kinds:

@@

\begin{compactitem}[---]
\item {\it Constructed:} Expressions can be constructed by invoking the
  \c{quote}, \c{enquote}, \c{expression}. Function arguments are passed as
  promises, and \c{substitute} is used to retrieve the source expression
  associated with the promise.
  \item {\it Reflection:} This group corresponds to uses of the \c{match.call}
  function to reflectively capture the expression that invoked the current
  function.
\item {\it String:} Finally, expressions can be created from strings by invoking
   \c{parse}, \c{str2expression}, or \c{str2lang}.
\end{compactitem}

@@

From the \packageNbrunsRnd programs we capture \SEAllRnd side effects from
\SEAllCallsRnd \eval calls (\SEUserCallsToAllRatio) in \SEAllSites sites
(\SEUserSitesToAllRatio). Again, the challenge is to remove the accidental
side effects caused by the R virtual machine implementation and are not
related to the user code. For example, the \c{.Random.seed} variable, which
contains the state of the random number generator, is saved and restored from
and to the global environment every time the user calls one of the base
routines that use random numbers.
%
% \begin{wrapfigure}{r}{5cm} \hspace*{-20mm} \centering
% \includegraphics[width=90mm]{se-types.pdf} \caption{Distribution of R types
% in target environments for side effects.} \label{fig:se-types}
% \end{wrapfigure}
%
Removing them leaves us with \SEUserRnd side effects (\SEUserRatio) from
\SEUserCallsRnd \eval calls (\SEUserCallsRatio) in \SEUserSites sites
(\SEUserSitesRatio). These sites are part of \SEUserFunctions R functions
(\SEUserFunctionsToAllRatio of all functions doing \eval) in \SEUserPackages
packages. From these functions, \SEFunsNighty are responsible for 90\% of side
effects. Half of the side-effects come just from three functions:
\c{plyr::allocate\_column} (allocates space for a new data frame column),
\c{withr::execute\_handlers} (executes deferred expressions), and
\c{foreach::doSEQ} (executes an expression on each element in a collection,
possibly in parallel). Most of the \eval sites (\SESitesInEnvirRatio) do side
effects in the environment specified by the \c{envir} parameter (\cf
Section~\ref{sec:eval-in-r}), \SESitesNotInEnvirRatio modifies other
environments, and finally, \SESitesBothEnvirRatio does both.
Table~\ref{tab:se-env} shows the class of the environment where \eval side
effects happen (\emph{target environment}). It follows the classification
described in Section~\ref{sec:env} with two extra classes: \emph{Object},
which represents a side-effect that happens in S4 or R6 object environment, and
\emph{Local} that corresponds to a side effect in local variables of the \eval
caller. The table shows data for both \eval sites and functions. For a
function to have a given target environment class, all of its \eval sites that
do side effects must have the same target environment class The majority of the
\eval sites (\SESitesInOneClass) do all side effects consistently in one
environment class. The same happens on the function level. Almost half of the
sites (over a third of the functions) do side effects in either \emph{Local} or
\emph{Object} environments. This gives a ray of hope for the compiler. Even
though it is possible to do anything anywhere, the data suggest that most
side-effects are sane.

>>>>

Contrary to the Javascript \eval, which does not precise the scope in which to evaluate,the second argument of \eval in R, \c{envir}, specifically is dedicated to that. 
The environment in which evaluation happens determines what is visible to the
computation started by \eval and the potential reach of its side-effects. The
second argument to \eval specifies that environment. Environments of \eval can be
classified into the following four kinds:

@@

\begin{compactitem}[---]
\item {\it Constructed:} Expressions can be constructed by invoking the
  \c{quote}, \c{enquote}, \c{expression} functions. Function arguments are passed as
  promises, and \c{substitute} is used to retrieve the source expression
  associated with the promise.
  \item {\it Reflection:} This group corresponds to uses of the \c{match.call}
  function to reflectively capture the expression that invoked the current
  function.
\item {\it String:} Finally, expressions can be created from strings by invoking
   \c{parse}, \c{str2expression}, or \c{str2lang}.
\end{compactitem}

@@

From the \packageNbrunsRnd programs, we capture \SEAllRnd side effects from
\SEAllCallsRnd \eval calls (\SEUserCallsToAllRatio) in \SEAllSites sites
(\SEUserSitesToAllRatio). Again, the challenge is to remove the accidental
side effects caused by the R virtual machine implementation that are not
related to the user code. For example, the \c{.Random.seed} variable, which
contains the state of the random number generator, is saved and restored from
and to the global environment every time the user calls one of the base
routines that use random numbers.
%
% \begin{wrapfigure}{r}{5cm} \hspace*{-20mm} \centering
% \includegraphics[width=90mm]{se-types.pdf} \caption{Distribution of R types
% in target environments for side effects.} \label{fig:se-types}
% \end{wrapfigure}
%
Removing them leaves us with \SEUserRnd side effects (\SEUserRatio) from
\SEUserCallsRnd \eval calls (\SEUserCallsRatio) in \SEUserSites sites
(\SEUserSitesRatio). These sites are part of \SEUserFunctions R functions
(\SEUserFunctionsToAllRatio of all functions doing \eval) in \SEUserPackages
packages. From these functions, \SEFunsNighty are responsible for 90\% of side
effects. Half of the side-effects come just from three functions:
\c{plyr::allocate\_column} (allocates space for a new data frame column),
\c{withr::execute\_handlers} (executes deferred expressions), and
\c{foreach::doSEQ} (executes an expression on each element in a collection,
possibly in parallel). Most of the \eval sites (\SESitesInEnvirRatio) do side
effects in the environment specified by the \c{envir} parameter (\cf
Section~\ref{sec:eval-in-r}), \SESitesNotInEnvirRatio modifies other
environments, and finally, \SESitesBothEnvirRatio does both.
Table~\ref{tab:se-env} shows the class of the environment where \eval side
effects happen (\emph{target environment}). It follows the classification
described in Section~\ref{sec:env} with two extra classes: \emph{Object},
which represents a side-effect that happens in S4 or R6 object environments, and
\emph{Local} that corresponds to a side effect in local variables of the \eval
caller (offset 0 in Section~\ref{sec:env}).  The table shows data for both \eval sites and functions. For a
function to have a given target environment class, all of its \eval sites that
do side effects must have the same target environment class The majority of the
\eval sites (\SESitesInOneClass) do all side effects consistently in one
environment class. The same happens at the function level. Almost half of the
sites (over a third of the functions) do side effects in either \emph{Local} or
\emph{Object} environments. This gives a ray of hope for the compiler. Even
though it is possible to do anything anywhere, the data suggest that most
side-effects are sane.
